---
title: "Seattle's Cycle Sharing Analysis"
author: "Hannah Xiao Si Laws, Sammy Pshyk, Megan Gillis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



We find the orignal information [here](https://proquestionasker.github.io/projects/2016/11/10/bicyclesseattle/).



```{r, echo = FALSE, warning = FALSE, message=FALSE}
#loading data

# For data manipulation and tidying
library(dplyr)
library(lubridate)
library(tidyr)

# For mapping
library(ggmap)
library(mapproj)

# For data visualizations
library(ggplot2)

# For modeling and machine learning
library(caret)
library(lubridate)

station <- read.csv(file = "station.csv", header = TRUE, 
                    stringsAsFactors = FALSE)

trip <- read.csv(file = "trip.csv", header = TRUE, 
                 stringsAsFactors = FALSE)

weather <- read.csv(file = "weather.csv", header = TRUE, 
                    stringsAsFactors = FALSE)
```
## **Introduction**
This report and analysis on Seattle's bicycle sharing trends. The data includes weather reports for the area, the station locations, and trips taken by cycle riders. Future explorations could include the addition of other forms of public transportation for the Seatle area to better understand commuter travel trends.

## **Station Locations**
Lets take a look at where the rental stations are located in Seattle!

```{r, label = "stationLocs", echo=FALSE}
#For show (the graphs , THE EVI-DANCE) Formating the dates
station$install_date <- mdy(station$install_date)


# How many times were new stations installed?
station %>% summarise(n_distinct(install_date))
# How many stations were installed on each date
station %>% group_by(install_date) %>% summarise(count = n()) %>% 
  arrange(install_date)



station %>% summarise(n_distinct(station_id))

station_locs <- station %>% group_by(station_id) %>% select(1:4, -2)

mymap <- get_map(location = c(lat = 47.60, lon = -122.35), maptype = "roadmap", zoom = 12)

# Plot a single point for each Station ID
ggmap(mymap) + geom_point(aes(x = long, y = lat), data = station_locs, 
                          alpha = 1, color = "darkred", size = 2)
```


```{r, label = "closeUp", fig.cap= "Stations map Zoomed In", warning=FALSE, message=FALSE}
closeUp <- get_map(location = "Seattle", maptype = "roadmap", zoom = 13)

# Plot a single point for each Station ID
ggmap(closeUp) + geom_point(aes(x = long, y = lat), data = station_locs, 
                            alpha = 1, color = "darkred", size = 3)
```


Look at all those stations!


## Histogram of Number of Bikes per Station

```{r, label = "nbs", fig.cap = "Current Dock Count", echo = FALSE, message = FALSE}
#histogram showing current dock count
ggplot(data = station, aes (x = current_dockcount)) +
  geom_bar(fill = "pink", color = "black") +
  theme_classic() +
  labs( x = "Number of bikes per station", y= "count", title = "Current Dock Count")
```
Here is a histogram of the number of bikes per station. 





```{r, label = "bikesPerStation", fig.cap="Change in Number of Bike Docks Per Station", warning = FALSE, message=FALSE, echo=FALSE}
dock_change <- station %>% 
               group_by(station_id) %>% 
               select(station_id, long, lat, ends_with("dockcount")) %>% 
               mutate(dock_change = current_dockcount -install_dockcount)


#attempt to recreate map plot thing

closeUp <- get_map(location = "Seattle", maptype = "roadmap", zoom = 13)

ggmap(closeUp) + 
  geom_point(aes(x = long, y = lat, size = factor(dock_change), color = factor(dock_change)), data = dock_change, alpha = 0.8) + 
  guides(color = guide_legend(title = "change"), size = guide_legend(title = "change")) +
  scale_size_manual(values = 10:1)
```
Here is a map of how many times dock changed occured per station. 

```{r, label = "currentStationSize", fig.cap="Current Station Size", warning = FALSE, message = FALSE, echo=FALSE}
currDockSize <- station %>% 
               group_by(station_id) %>% 
               select(station_id, long, lat, ends_with("dockcount")) %>% 
               mutate(currDockSize = current_dockcount)

closeUp <- get_map(location = "Seattle", maptype = "roadmap", zoom = 13)

ggmap(closeUp) + 
  geom_point(aes(x = long, y = lat, size = factor(currDockSize), color = factor(currDockSize)), data = currDockSize, alpha = 0.8) + 
  guides(color = guide_legend(title = "change"), size = guide_legend(title = "change")) +
  scale_size_manual(values = 10:1)
```


11 stations lost bike docks, 39 docks stayed the same, 8 stations gained docks.


```{r, label = "visualRidesPerDay", fig.cap="Time to visualize the number of rides per day", message = FALSE, echo=FALSE}
# Make the start and stop dates into POSIXct objects
trip_2 <- trip %>% mutate(start_dt = mdy_hm(starttime), stop_dt = mdy_hm(stoptime))

trip_2 <- trip_2 %>% mutate(start_date = paste(month(start_dt), 
                                               day(start_dt), year(start_dt), sep = "/"))
trip_2$start_date <- mdy(trip_2$start_date)

trip_2 <- trip_2 %>% mutate(stop_date = paste(month(stop_dt), 
                                              day(stop_dt), year(stop_dt), sep = "/"))
trip_2$stop_date <- mdy(trip_2$stop_date)


# Recode the dates
trip_2 %>% 
  
  group_by(start_date) %>%
  
  summarize(N = n()) %>%
  
  ggplot(aes(x = start_date, y = N)) + 
  
  geom_line() + 
  
  labs(x = "Date", y = "Number of trips per day") + 
  
  theme_bw() + geom_smooth()


```

#Plotting trips per month (by season, minus summer)
```{r, label = "monthSeason", fig.cap="Trips Per Month (By Season)", echo=FALSE}
start_date_ym <- trip_2 %>% 
  mutate(ym = paste(year(start_date), 
                    month(start_date), sep = "/"), Season = ifelse(ym %in% c("2014/10", "2014/11"), "Fall",
                                                            ifelse(ym %in% c("2014/12", "2015/1", "2015/2"), "Winter",
                                                            ifelse(ym %in% c("2015/3", "2015/4", "2015/5"), "Spring", "Summer"))))

start_date_ym %>%
  group_by(ym, Season) %>%
  summarize(N=n()) %>%
  ggplot(aes(x = ym, y = N, color = Season, fill = Season, group = Season)) + 
  geom_point() +
  geom_line(group = 1) + 
  labs( x = "Date", y = "Number of Trips(per month)")
```



# Convert Trip Duration from Seconds to Minutes
```{r, label = "minutesSeason", fig.cap="Average Trip Duration", echo=FALSE}
Trip_Duration_Month <- start_date_ym %>% 
  mutate(trip_duration_min = tripduration/60) %>% 
  group_by(ym) %>% 
  select(ym, trip_duration_min, Season) %>% 
  summarise(Avg = mean(trip_duration_min), 
            sd = sd(trip_duration_min)) %>% 
  mutate(se = sd/sqrt(n()), Season = ifelse(ym %in% c("2014/10", "2014/11"), "Fall",
                                     ifelse(ym %in% c("2014/12", "2015/1", "2015/2"), "Winter",
                                     ifelse(ym %in% c("2015/3", "2015/4", "2015/5"), "Spring", "Summer"))))

Trip_Duration_Month %>%
  ggplot(aes(x = ym, y = Avg, col = Season, fill = Season, group = Season)) + 
  geom_point()+
  geom_line(aes(group = 1)) + 
  labs(x = "Date" , y = "Duration of Average Trip (minutes)") + 
  theme_bw() + 
  geom_errorbar(aes(ymin = Avg - se, ymax = Avg + se))

```


##Number of Trips by Day of Week
```{r, label = "dayOfWeekSeasons", echo = FALSE, warning=FALSE, message=FALSE}
 trip_2$wd <- wday(trip_2$start_date, label = TRUE)
 tripduration_m <- (trip_2$tripduration)/60
 trip22<-trip_2
 trip22<-cbind(trip22, tripduration_m)
 ym <- format(trip22$start_date, "%Y/%m")
 trip22<- cbind(trip22,ym)
 
 trip22 <-mutate(trip22, Season = ifelse(ym %in% c("2014/10", "2014/11"), "Fall",
                                      ifelse(ym %in% c("2014/12", "2015/1", "2015/2"), "Winter",
                                      ifelse(ym %in% c("2015/3", "2015/4", "2015/5"), "Spring", "Summer"))))%>%
   group_by(Season)
 

trip22 %>%
  group_by(wd, Season) %>%
  summarize(N = n())%>%
  ggplot(aes(x = wd, y = N, col = Season, fill = Season, group = Season)) + 
  geom_point() + 
  geom_line() + 
  theme_bw() + 
  labs(x = "Day of the Week", y= "Number of Trips", title = "Trips Per Day")


```

##Number of Trips Per Time of Day

```{r, label = "timeOfDaySeasons3", message = FALSE, warnings = FALSE, echo = FALSE}

trip22 %>%
  group_by(Hour = hour(start_dt), Season, wd) %>%
  summarize(N = n()) %>%
 ggplot(aes(x = Hour, y= N, color = Season, group = Season)) +
  geom_point() +
  geom_line() + 
  facet_wrap(~wd) + 
  theme_bw() +
  labs(x = "Hour of Day", y = "Number of Trips")

```

##Number of Trips by Member Type
```{r, label = "timeOfDaySeasons2", message = FALSE, warnings = FALSE, echo = FALSE}

trip22 %>%
  filter(usertype !="") %>%
  group_by(Season, usertype, ym) %>%
  summarize(N = n()) %>%
  ggplot(aes(x = ym, y = N, color = Season, group = usertype, fill = usertype)) + 
  geom_point(aes(shape = usertype), size = 3) + 
  geom_line() + 
  theme_bw() + 
  labs(x = "Date", y= "Total Number of Trips Per Month")

```

##Trip Duration by Member Type
```{r, message = FALSE, warnings = FALSE, echo = FALSE}

#Frequency histogram
start_date_ym %>%
  filter(usertype !="")%>%
  ggplot(aes(x = tripduration, fill = usertype))+
  geom_histogram() +
  facet_grid(.~ usertype) + 
  scale_fill_manual(values = c("pink", "light blue")) +
  theme_bw() +
  guides(fill = FALSE) +
  geom_vline(aes(xintercept = quantile(tripduration, prob = 0.75)), linetype = "dashed")



```

##Member Demographics
```{r, label = "timeOfDaySeasons1", message = FALSE, warnings = FALSE, echo = FALSE}

trip_2$usertype <- as.factor(trip_2$usertype)
trip_age <- trip_2 %>% mutate(age = year(start_dt) - birthyear)

hist(trip_age$age, main = "Member Age", xlab = "Number of Riders", 
    col = "red", breaks = 25)


```{r, eval=FALSE, label = "timeOfDaySeasons", fig.cap="Number of Trips by Time of Day", message = FALSE, echo=FALSE}
#Number of Trips Per Time of Day

start_date_ym%>%
  group_by(starttime,wd, Season)%>%
  summarize(N = n()) %>%
  ggplot(aes(x = starttime, y= N, col = Season, fill = Season, group = Season)) +
  geom_point() +
  geom_line() + 
  facet_grid(.~wd) + 
  labs(x = "Hour of Day", y = "Number of Trips")


```


```{r, eval = FALSE, echo = FALSE, message = FALSE}
#Trip Duration by Member Type
trip_cost <- trip_2 %>% mutate(cost = ifelse(usertype == "Member" & tripduration_m <= 45, 0, 
                                      ifelse(usertype == "Member" & tripduration_m >45 & tripduration_m <= 75, 2, 
                                      ifelse(usertype == "Member" & tripduration_m > 75, (2 + 2 * ((tripduration_m - 75)/30 #*ALERT NEEDS TO BE CHANGED*)),
                                      ifelse(usertype == "Short-Term Pass Holder" & tripduration_m <= 30, 0, 
                                      ifelse(usertype == "Short-Term Pass Holder" & tripduration_m > 30 & tripduration_m < 60, 2, 
                                      ifelse(usertype == "Short-Term Pass Holder" & tripduration_m > 60, 
                                             (2 + 5 * ((tripduration_m - 60)/30)), NA #any extra overage is $5, needs to be fixed)))))))

                                        
#attempt to make member vs. short term pass holder side by side graph
ggplot(trip_cost, aes(x = usertype)) + facet_grid(cost~.)

# start_date_ym <- start_date_ym %>% mutate(cost = ifelse(usertype == "Member" & tripduration_m <= 45,0, ifelse(usertype= "Member" & tripduration_m > 45 & tripduration_m <= 75, 2, ifelse(usertype = "Member" & tripduration_m > 75, (2+2 * ceiling(tripduration_m -75)/30, ifelse(usertype = "Short-Term Passholder" & tripduration_m <= 30,0 ifelse(usertype == "Short-Term Passholder"))))))
```


```{r, echo = FALSE, message = FALSE}
#Members Demographic
trip_2$usertype <- as.factor(trip_2$usertype)
trip_age <- trip_2 %>% mutate(age = year(start_dt) - birthyear)

hist(trip_age$age, main = "Member Age", xlab = "Number of Riders", 
     col = "#56B4E9", breaks = 25)
```






```{r, eval = FALSE, echo = FALSE, message = FALSE}
#weather data stuff
str(weather)
# Adjusting the Date Variable
weather$Date <- mdy(weather$Date)

# Adjusting the Events Variable
weather$Events <- as.factor(weather$Events)

#Let’s combine a few of these things that seem to represent the same event.
weather$Events <- gsub("Fog , Rain|Fog-Rain", "Fog-Rain", weather$Events)
weather$Events <- gsub("Rain , Snow|Rain-Snow", "Rain-Snow", 
                       weather$Events)
weather$Events <- gsub("Rain , Thunderstorm|Rain-Thunderstorm", 
                       "Rain-TS", weather$Events)

weather$Events <- as.factor(weather$Events)


weather$Max_Gust_Speed_MPH <- gsub("-", 0, weather$Max_Gust_Speed_MPH)

weather$Max_Gust_Speed_MPH <- as.numeric(weather$Max_Gust_Speed_MPH)



weather[which(is.na(weather$Mean_Temperature_F)), 1]


weather[490, "Mean_Temperature_F"] <- "50"

weather$Events <- gsub("^$", "Other", weather$Events)
weather$Events <- as.factor(weather$Events)



#weather and trip data combo
# Make a copy of the data frame
trip_3 <- trip_2

# Change column name in trip_3 to match weather dataset
trip_3$Date <- trip_3$start_date

# Left join the trip and weather dataframes by date.
trip_weather <- left_join(trip_3, weather, by = "Date")
```




```{r, eval = FALSE, echo = FALSE}
#Analysis

```



```{r, eval = FALSE, echo = FALSE}

#our direction

# looking into wind speed and trips
# blah blah blah
```











